---
title: "Homework 9"
author: "STAT 601"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F,warning=F,echo=F,fig_height=10,fig_width=7,cache = F)
```

Please do the following problems from the text book R Handbook and discuss your analysis with the rest of the class.  Remember to have an annotated RMD file that will allow the rest of us to see what you are doing!


1. Following up with the Beat the Blues data from the video (package HSAUR3) do the following
```{r}
#Setup
usePackage <- function(p) {
    if (!is.element(p, installed.packages()[,1]))
        install.packages(p, dep = TRUE)
    require(p, character.only = TRUE)
}

usePackage("HSAUR3")
usePackage("gee")
usePackage("lme4")
usePackage("Matrix")
usePackage("multcomp")
usePackage("gridExtra")
usePackage("ggplot2")


```


      a. Construct boxplots to compare the factor variable \textbf{drug} in an analogous way to how we constructed boxplots in the video for the treatment variable. Discuss the results. 
```{r}
head(BtheB)
class(BtheB)

layout(matrix(1:2, nrow = 1))  
ylim <- range(BtheB[,grep("bdi", names(BtheB))],na.rm = TRUE)  
tau <- subset(BtheB, treatment == "TAU")[,grep("bdi", names(BtheB))]
boxplot(tau, main = "Treated as Usual", ylab = "BDI",
        xlab = "Time (in months)", names = c(0, 2, 3, 5, 8),
        ylim = ylim)
btheb <- subset(BtheB, treatment == "BtheB")[,grep("bdi", names(BtheB))]
boxplot(btheb, main = "Beat the Blues", ylab = "BDI", xlab = "Time (in months)", names = c(0, 2, 3, 5, 8),ylim = ylim) 

layout(matrix(1:2, nrow = 1))  
ylim <- range(BtheB[,grep("bdi", names(BtheB))],na.rm = TRUE)  
tau <- subset(BtheB, length == ">6m")[,grep("bdi", names(BtheB))]
boxplot(tau, main = "Time Depressed >6m", ylab = "BDI",
        xlab = "Time (in months)", names = c(0, 2, 3, 5, 8),
        ylim = ylim)
btheb <- subset(BtheB, length == "<6m")[,grep("bdi", names(BtheB))]
boxplot(btheb, main = "Time Depressed <6m", ylab = "BDI", xlab = "Time (in months)", names = c(0, 2, 3, 5, 8),ylim = ylim) 


layout(matrix(1:2, nrow = 1))  
ylim <- range(BtheB[,grep("bdi", names(BtheB))],na.rm = TRUE)  
tau <- subset(BtheB, drug == "Yes")[,grep("bdi", names(BtheB))]
boxplot(tau, main = "Antidepressants = Yes", ylab = "BDI",
        xlab = "Time (in months)", names = c(0, 2, 3, 5, 8),
        ylim = ylim)
btheb <- subset(BtheB, drug == "No")[,grep("bdi", names(BtheB))]
boxplot(btheb, main = "Antidepressants = No", ylab = "BDI", xlab = "Time (in months)", names = c(0, 2, 3, 5, 8),ylim = ylim) 


unique(BtheB$drug)

```
*I created new box plots showing the other factors of the analysis which include whether the patient took anti-depressant drugs, and whether the patient currently had an episode of depression more or less than 6 months. Results are similar to the original boxplot comparing the treatments of the patients although there are more outliers. Also, I'm noticing more of a decrease over time in the Beck Depression Inventory II measure when applying treatments on patients who are in a depression episode less than 6 months.*

      b. Repeat (a) for the \textbf{length} variable. Discuss the results.
      
  *see part a*

      c. Use the \textit{lm} function to fit a model to the Beat the Blues data that assumes that the repeated measurements are independent. Compare the results to those from fitting the random intercept model \textit{BtheB\_lmer1} from the video.
      
    ```{r}
 #Change the dataset to long form to allow for other analysis
BtheB$subject <- factor(rownames(BtheB))
nobs <- nrow(BtheB)
BtheB_long <- reshape(BtheB, idvar = "subject", varying = c("bdi.2m", "bdi.3m", "bdi.5m", "bdi.8m"), direction = "long")
BtheB_long$time <- rep(c(2, 3, 5, 8), rep(nobs, 4)) 

#fit lmer model
BtheB_lmer1 <- lmer(bdi ~ bdi.pre + time + treatment + drug +   length + (1 | subject), data = BtheB_long, REML = FALSE, na.action = na.omit)  
BtheB_lmer2 <- lmer(bdi ~ bdi.pre + time + treatment + drug +   length + (time | subject), data = BtheB_long, REML = FALSE, na.action = na.omit)  
anova(BtheB_lmer1, BtheB_lmer2) 

# layout(matrix(1:2, ncol = 2))  
# qint <- ranef(BtheB_lmer1)$subject[["(Intercept)"]]  
# qres <- residuals(BtheB_lmer1)  
# qqnorm(qint, ylab = "Estimated random intercepts", 
#        xlim = c(-3, 3), 
#        ylim = c(-20, 20), main = "Random intercepts")  
# qqline(qint)  
# qqnorm(qres, xlim = c(-3, 3), ylim = c(-20, 20), ylab = "Estimated residuals", main = "Residuals")
# qqline(qres) 


#fit lm model
BtheB_lm1 <- lm(bdi ~ bdi.pre + time + treatment + drug  + length, data = BtheB_long, na.action = na.omit)  
# BtheB_lm2 <- lm(bdi ~ bdi.pre + time + treatment + drug  + length, data = BtheB_long,  na.action = na.omit)
# anova(BtheB_lm1, BtheB_lm2) 


cftest(BtheB_lmer1)
cftest(BtheB_lmer2)
summary(BtheB_lm1)

# layout(matrix(1:2, ncol = 2))  
# qint <- ranef(BtheB_lm1)$subject[["(Intercept)"]]  
# qres <- residuals(BtheB_lm1)  
# qqnorm(qint, ylab = "Estimated  intercepts", 
#        xlim = c(-3, 3), 
#        ylim = c(-20, 20), main = " intercepts")  
# qqline(qint)  
# qqnorm(qres, xlim = c(-3, 3), ylim = c(-20, 20), ylab = "Estimated residuals", main = "Residuals")
# qqline(qres)




```
   *When comparing the random intercept models with a linear model of the data, the linear model, along with bdi.pre, and time, find that whether the Beat the Blues Treatment was used and whether the patient was already medicated with anti-depressants are significant factors for determining patient bdi.*   

      d. Investigate and discuss whether there is any evidence of an interaction between treatment and time for the Beat the Blues data.
```{r}
#Check for an interaction between treatment and time via linear model
BtheB_lm_1d <- lm(time ~ treatment, data = BtheB_long, na.action = na.omit)  
summary(BtheB_lm_1d)

BtheB_lm_1d2 <- lm(bdi ~ bdi.pre + time * treatment + drug  + length, data = BtheB_long, na.action = na.omit)  
summary(BtheB_lm_1d2)
#The linear model does not show an interaction between treatment and time

```
*When fitted to a linear model, it seems like there is do interaction between treatment and time.*      

      e. Construct a plot of the mean profiles of both treatment groups in the Beat the Blues data, showing also standard deviation bars at each time point.
      
```{r}

 # library
library(ggplot2)

#Split dataset between Treatments
BtheB_TAU <- subset(BtheB, treatment == "TAU")
BtheB_BTB <- subset(BtheB, treatment != "TAU")
 
# create a dataset
months <- c(rep("pre" , 2) , rep("2" , 2) , rep("3" , 2) , rep("5" , 2), rep("8" , 2) )
measure <- rep(c("mean" , "sd"), 5)
bdi_tau <- c(mean(BtheB_TAU$bdi.pre, na.rm = TRUE),sd(BtheB_TAU$bdi.pre, na.rm = TRUE),mean(BtheB_TAU$bdi.2m, na.rm = TRUE),sd(BtheB_TAU$bdi.2m, na.rm = TRUE),mean(BtheB_TAU$bdi.3m,na.rm = TRUE),sd(BtheB_TAU$bdi.3m, na.rm = TRUE),mean(BtheB_TAU$bdi.5m, na.rm = TRUE),sd(BtheB_TAU$bdi.5m, na.rm = TRUE),mean(BtheB_TAU$bdi.8m, na.rm = TRUE),sd(BtheB_TAU$bdi.8m, na.rm = TRUE))
bdi_btb <- c(mean(BtheB_BTB$bdi.pre, na.rm = TRUE),sd(BtheB_BTB$bdi.pre, na.rm = TRUE),mean(BtheB_BTB$bdi.2m, na.rm = TRUE),sd(BtheB_BTB$bdi.2m, na.rm = TRUE),mean(BtheB_BTB$bdi.3m,na.rm = TRUE),sd(BtheB_BTB$bdi.3m, na.rm = TRUE),mean(BtheB_BTB$bdi.5m, na.rm = TRUE),sd(BtheB_BTB$bdi.5m, na.rm = TRUE),mean(BtheB_BTB$bdi.8m, na.rm = TRUE),sd(BtheB_BTB$bdi.8m, na.rm = TRUE))
data <- data.frame(months,measure,bdi_tau,bdi_btb)
 
# Grouped
plot1 <- ggplot(data, aes(fill=measure, y=bdi_tau, x=months)) + 
    geom_bar(position="dodge", stat="identity") +
  ylim(min=0, max=25)
plot2 <- ggplot(data, aes(fill=measure, y=bdi_btb, x=months)) + 
    geom_bar(position="dodge", stat="identity") +
  ylim(min=0, max=25)


grid.arrange(plot1, plot2, ncol=2, nrow=1)


# create a dataset
months2 <- c(rep("pre" , 1) , rep("2" , 1) , rep("3" , 1) , rep("5" , 1), rep("8" , 1) )
measure2 <- rep(c("mean"), 5)
bdi_tau2 <- c(mean(BtheB_TAU$bdi.pre, na.rm = TRUE),mean(BtheB_TAU$bdi.2m, na.rm = TRUE),mean(BtheB_TAU$bdi.3m,na.rm = TRUE),mean(BtheB_TAU$bdi.5m, na.rm = TRUE),mean(BtheB_TAU$bdi.8m, na.rm = TRUE))
bdi_btb2 <- c(mean(BtheB_BTB$bdi.pre, na.rm = TRUE),mean(BtheB_BTB$bdi.2m, na.rm = TRUE),mean(BtheB_BTB$bdi.3m,na.rm = TRUE),mean(BtheB_BTB$bdi.5m, na.rm = TRUE),mean(BtheB_BTB$bdi.8m, na.rm = TRUE))
data2 <- data.frame(months,measure,bdi_tau,bdi_btb)


library(dplyr)
library(tidyr)

# build an adjusted londatasets
 #Change the dataset to long form to allow for other analysis
BtheB$subject <- factor(rownames(BtheB))
nobs <- nrow(BtheB)
BtheB_long2 <- reshape(BtheB, idvar = "subject", varying = c("bdi.pre","bdi.2m", "bdi.3m", "bdi.5m", "bdi.8m"), direction = "long")
BtheB_long2$time <- rep(c(0,2, 3, 5, 8), rep(nobs, 5)) 

BtheB_plot <- BtheB_long2 %>% 
    group_by(treatment, time) %>% 
    summarise(mean= mean(bdi, na.rm = TRUE), sd = sd(bdi, na.rm = TRUE))

# Corrected with actual standard deviation bars
ggplot(BtheB_plot, aes(time, mean, fill=treatment)) + 
  geom_bar(stat="identity", width = 0.7, position = position_dodge(width=0.7)) +
  geom_errorbar(aes(ymin=mean-sd, max=mean+sd), width=.2, position=position_dodge(width=0.7))


```
*These bar charts show that the means of the two treatments tend to be noticeably lower for patients treated with the beat the blues therapy compared to the normal treatment and the standard deviations tend to be noticeably lower for months 5 and 8.*
  
  
2. Consider the \textbf{phosphate} data from the package HSAUR3. This data shows the plasma inorganic phosphate levels for 33 subjects, 20 of whom are controls and 13 of whom have been classified as obese (Davis, 2002). Perform the following on this dataset
 
      a. Construct boxplots by group and discuss. 
```{r}
unique(phosphate$group)
head(phosphate)
nrow(phosphate)



layout(matrix(1:2, nrow = 1))  
ylim <- range(phosphate[,grep("t", names(phosphate))],na.rm = TRUE)  
control <- subset(phosphate, group == "control")[,grep("t", names(phosphate))]
boxplot(control, main = "Control", ylab = "Phosphate level",
        xlab = "Time (in hours)", names = c(0, .5, 1, 1.5, 2, 3, 4, 5),
        ylim = ylim)
obese <- subset(phosphate, group == "obese")[,grep("t", names(phosphate))]
boxplot(obese, main = "Obese", ylab = "Phosphate level", xlab = "Time (in hours)", names = c(0, .5, 1, 1.5, 2, 3, 4, 5),ylim = ylim) 
```
      *Results of the boxplots seem to differ between the control and obese groups, with the obese group having more outliers and a trend with a trough at 2 hours in compared to the control group which has a trough at 1.5 hours in.*
   
      b. Produce separate plots of the profiles of the individuals in each group.
```{r}

phosphate.df <- as.data.frame(phosphate)
phosphate.df$subject <- factor(rownames(phosphate.df))

phosphate_long <- reshape(phosphate.df, 
        direction = "long",
        varying = list(names(phosphate)[2:9]),
        v.names = "Phos_Level",
        idvar = c("subject", "group"),
        timevar = "hour",
        times = c(0, .5, 1, 1.5, 2, 3, 4, 5))

phosphate_long_2 <- reshape(phosphate.df, 
        direction = "long",
        varying = list(names(phosphate)[3:9]),
        v.names = "Phos_Level",
        idvar = c("subject", "group", "t0"),
        timevar = "hour",
        times = c(.5, 1, 1.5, 2, 3, 4, 5))

BtheB$subject <- factor(rownames(BtheB))
nobs <- nrow(BtheB)
BtheB_long <- reshape(BtheB, idvar = "subject", varying = c("bdi.2m", "bdi.3m", "bdi.5m", "bdi.8m"), direction = "long")
BtheB_long$time <- rep(c(2, 3, 5, 8), rep(nobs, 4)) 


phosphate_long_obese <- subset(phosphate_long, group == "control")
phosphate_long_control <- subset(phosphate_long, group != "control")


plot_2b1 <- ggplot(phosphate_long_control, aes(x = hour, y = Phos_Level, colour = subject)) + geom_line()  +
  ylim(min=1, max=7)  + labs(title = "Control Subjects", colour = "Subjects", y= "phosphate level")
plot_2b2 <- ggplot(phosphate_long_obese, aes(x = hour, y = Phos_Level, colour = subject)) + geom_line()  +
  ylim(min=1, max=7)  + labs(title = "Obese Subjects", colour = "Subjects", y= "phosphate level")


grid.arrange(plot_2b1, plot_2b2, ncol=2, nrow=1)
```
      
    
      c. Guided by how these plots fit, which linear mixed effects models do you think might be sensible? (Hint:                Discuss intercept and slope, intercept and interaction).

*After comparing the two types of linear mixed effects models, I'd say that the random intercept model would be more sensible to fit to this dataset due to the measures not really having a particularly noticeable pattern in variance other than some subjects having a high variance while others having a low variance. Time between when the measures were taken does not seem to be having a noticeable impact on variance.*
    
      d. Convert the data to long version and fit the model of your choice and discuss the results.
```{r}
head(phosphate_long_2)
#test random intercept model and fitted random intercept model for this dataset
phosphate_lmer1 <- lmer(Phos_Level ~ group + hour + t0 + (1 | subject), data = phosphate_long_2, REML = FALSE, na.action = na.omit)  
phosphate_lmer2 <- lmer(Phos_Level ~ group + hour + t0 + (hour | subject), data = phosphate_long_2, REML = FALSE, na.action = na.omit)  
anova(phosphate_lmer1, phosphate_lmer2) 

summary(phosphate_lmer1)
summary(phosphate_lmer2)
cftest(phosphate_lmer1)
cftest(phosphate_lmer2)
```
  *The random intercept model seems to be a better fit based on AIC and BIC than the fitted random intercept model. Furthermore, when determining phosphate levels, it seems like hour and initial phosphate levels have a significant effect on phosphate levels.*
   